Component response rate variation underlies the stability of complex systems

### 2019-03-03 notebook/arxiv/ms.tex
Tweaks

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) might be expected to decrease system stability
by introducing a new source of variation into the system and thereby
increasing \(\sigma\). Here I show why, despite higher \(\sigma\),
realistic complex systems (such that \(S\) is high but finite) in which
\(\gamma\) varies are actually more likely to be stable.

@@

My focus is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity, \(\sigma\sqrt{SC}\) (see Supplemental Information for
results simulated across \(\sigma\) and \(C\)). Overall, I show that
component response rate variation increases the upper bound of
complexity at which stability can be realistically observed, meaning
that highly complex systems are more likely than not to vary in their
component response rates, and for this variation to be critical for
system stability.

@@

In the above, \(\mathbf{n^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{n^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
stability
criterion\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below (see
also\textsuperscript{\protect\hyperlink{ref-Servan2018}{22}}),

@@

\hypertarget{ref-Gibbs2017}{}
17. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{arXiv} (2017).

@@

\begin{center}
\hypertarget{SIstart}{\section{Supplementary Information}\label{SIstart}}
\end{center}

@@

\textbf{This supplementary information supports the manuscript
``Component response rate variation drives stability in large complex
systems'' with all of the code required to recreate the analysis in the
main text, and with additional analyses to support its conclusions. All
text, code, and data underlying this manuscript are publicly available
on \href{https://github.com/bradduthie/RandomMatrixStability}{GitHub}
as part of the RandomMatrixStability package.}

The
\href{https://github.com/bradduthie/RandomMatrixStability}{RandomMatrixStability
package} includes all functions and tools for recreating the text, this
supplemental information, and running all code; additional documentation
is also provided for functions as part of the package. The
RandomMatrixStability package is available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub}; to
download it, the
\href{https://cran.r-project.org/web/packages/devtools/index.html}{\texttt{devtools}
library} is needed.

@@

While downloading this package is recommended, all relevant code is also
reproduced below with explanation, so it is possible to recreate all
analyses using only this Supplementary Information.

@@

\section{Supplementary Information table of
contents}\label{supplemental-information-table-of-contents}

\begin{itemize}
\tightlist
\item
  \protect\hyperlink{moregamma}{Further explanation of \(\gamma\)}
\item
  \protect\hyperlink{Fig1}{Code and simulations underlying Fig. 1}
\item
  \protect\hyperlink{Fig2}{Code and simulations underlying Fig. 2}
\item
  \protect\hyperlink{IncrS}{Stability across increasing \(S\)}
\item
  \protect\hyperlink{ecological}{Stability of ecological networks}

  \begin{itemize}
  \tightlist
  \item
    \protect\hyperlink{competition}{Competitor networks}
  \item
    \protect\hyperlink{mutualism}{Mutualist networks}
  \item
    \protect\hyperlink{pred-prey}{Predator-prey networks}
  \end{itemize}
\item
  \protect\hyperlink{connectance}{Different connectance (C) values}

  \begin{itemize}
  \tightlist
  \item
    \protect\hyperlink{connect3}{C = 0.3}
  \item
    \protect\hyperlink{connect5}{C = 0.5}
  \item
    \protect\hyperlink{connect7}{C = 0.7}
  \item
    \protect\hyperlink{connect9}{C = 0.9}
  \end{itemize}
\item
  \protect\hyperlink{gam_dist}{Different distributions of \(\gamma\)}
\item
  \protect\hyperlink{ga}{Genetic algorithm}
\item
  \protect\hyperlink{Feasibility}{Feasibility of complex systems}
\end{itemize}

\hypertarget{moregamma}{\section{\texorpdfstring{Further explanation of
\(\gamma\)}{Further explanation of \textbackslash{}gamma}}\label{moregamma}}

In a synthesis of eco-evolutionary feedbacks on community stability,
Patel et al. model a system that includes a vector of potentially
changing species densities (\(\mathbf{N}\)) and a vector of potentially
evolving traits
(\(\mathbf{x}\))\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}}.
For any species \(i\) or trait \(j\), change in species density
(\(N_{i}\)) or trait value (\(x_{j}\)) with time (\(t\)) is a function
of the vectors \(\mathbf{N}\) and \(\mathbf{x}\),

In the above, \(f_{i}\) and \(g_{j}\) are functions that define the
effects of all species densities and trait values on the density of a
species \(i\) and the value of trait \(j\), respectively. Patel et al.
were interested in stability when the evolution of traits was relatively
slow or fast in comparison with the change in species
densities\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}}, and
this is modulated in the above by the scalar \(\epsilon\). The value of
\(\epsilon\) thereby determines the timescale separation between ecology
and evolution, with high \(\epsilon\) modelling relatively fast
evolution and low \(\epsilon\) modelling relative slow
evolution\textsuperscript{\protect\hyperlink{ref-Patel2018}{1}}.

I use the same principle that Patel et al. use to modulate the relative
rate of evolution to modulate rates of component responses for \(S\)
components. Following
May\textsuperscript{\protect\hyperlink{ref-May1972}{2},\protect\hyperlink{ref-May1973}{3}},
the value of a component \(i\) at time \(t\) (\(v_{i}(t)\)) is affected
by the value of \(j\) (\(v_{j}(t)\)) and \(j\)'s marginal effect on
\(i\) (\(m_{ij}\)), and by \(i\)'s response rate (\(\gamma_{i}\)),

\[\frac{dv_{i}(t)}{dt} = \gamma_{i} \sum_{j=1}^{S}m_{ij}v_{j}(t).\]

In matrix notation\textsuperscript{\protect\hyperlink{ref-May1973}{3}},

\[\frac{d\mathbf{v}(t)}{dt} = \mathbf{\gamma} \mathbf{M}\mathbf{v}(t).\]

In the above, \(\mathbf{\gamma}\) is a diagonal matrix in which elements
correspond to individual component response rates. Therefore,
\(\mathbf{\gamma} \mathbf{M}\) modulates the values of components and
can be analysed using the techniques of
May\textsuperscript{\protect\hyperlink{ref-May1972}{2},\protect\hyperlink{ref-May1973}{3}}.

\hypertarget{Fig1}{\section{Code and simulations underlying Fig.
1}\label{Fig1}}

The sample \(M\) used for the eigenvalue distributions in Fig. 1 of the
text is available on
\href{https://github.com/bradduthie/RandomMatrixStability/tree/master/notebook/sim_results/bi_gamma}{GitHub},
and was produced by running the following function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{find_bgamma <-}\StringTok{ }\NormalTok{function(}\DataTypeTok{S =} \DecValTok{200}\NormalTok{, }\DataTypeTok{C =} \FloatTok{0.05}\NormalTok{, }\DataTypeTok{Osd =} \FloatTok{0.4}\NormalTok{, }\DataTypeTok{iters =} \DecValTok{10000}\NormalTok{)\{}
    \NormalTok{while(iters >}\StringTok{ }\DecValTok{0}\NormalTok{)\{}
        \NormalTok{A_dat  <-}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(}\DataTypeTok{n =} \NormalTok{S *}\StringTok{ }\NormalTok{S, }\DataTypeTok{mean =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sd =} \NormalTok{Osd);}
        \NormalTok{A_mat  <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \NormalTok{A_dat, }\DataTypeTok{nrow =} \NormalTok{S);}
        \NormalTok{C_dat  <-}\StringTok{ }\KeywordTok{rbinom}\NormalTok{(}\DataTypeTok{n =} \NormalTok{S *}\StringTok{ }\NormalTok{S, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{, }\DataTypeTok{prob =} \NormalTok{C);}
        \NormalTok{C_mat  <-}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DataTypeTok{data =} \NormalTok{C_dat, }\DataTypeTok{nrow =} \NormalTok{S, }\DataTypeTok{ncol =} \NormalTok{S);}
        \NormalTok{A_mat  <-}\StringTok{ }\NormalTok{A_mat *}\StringTok{ }\NormalTok{C_mat;}
        \NormalTok{gammas <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\KeywordTok{rep}\NormalTok{(}\FloatTok{1.95}\NormalTok{, S/}\DecValTok{2}\NormalTok{), }\KeywordTok{rep}\NormalTok{(}\FloatTok{0.05}\NormalTok{, S/}\DecValTok{2}\NormalTok{))}
        \NormalTok{mu_gam <-}\StringTok{ }\KeywordTok{mean}\NormalTok{(gammas);}
        \KeywordTok{diag}\NormalTok{(A_mat) <-}\StringTok{ }\NormalTok{-}\DecValTok{1}\NormalTok{;}
        \NormalTok{A1     <-}\StringTok{ }\NormalTok{gammas *}\StringTok{ }\NormalTok{A_mat;}
        \NormalTok{A0     <-}\StringTok{ }\NormalTok{mu_gam *}\StringTok{ }\NormalTok{A_mat;}
        \NormalTok{A0_e   <-}\StringTok{ }\KeywordTok{eigen}\NormalTok{(A0)$values;}
        \NormalTok{A0_r   <-}\StringTok{ }\KeywordTok{Re}\NormalTok{(A0_e);}
        \NormalTok{A0_i   <-}\StringTok{ }\KeywordTok{Im}\NormalTok{(A0_e);}
        \NormalTok{A1_e   <-}\StringTok{ }\KeywordTok{eigen}\NormalTok{(A1)$values;}
        \NormalTok{A1_r   <-}\StringTok{ }\KeywordTok{Re}\NormalTok{(A1_e);}
        \NormalTok{A1_i   <-}\StringTok{ }\KeywordTok{Im}\NormalTok{(A1_e);}
        \NormalTok{if(}\KeywordTok{max}\NormalTok{(A0_r) >=}\StringTok{ }\DecValTok{0} \NormalTok{&}\StringTok{ }\KeywordTok{max}\NormalTok{(A1_r) <}\StringTok{ }\DecValTok{0}\NormalTok{)\{}
            \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{A0 =} \NormalTok{A0, }\DataTypeTok{A1 =} \NormalTok{A1));}
            \NormalTok{break;}
        \NormalTok{\}}
        \KeywordTok{print}\NormalTok{(iters);}
        \NormalTok{iters <-}\StringTok{ }\NormalTok{iters -}\StringTok{ }\DecValTok{1}\NormalTok{;}
    \NormalTok{\}}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

The above \texttt{find\_bgamma} function terminates when a matrix \(M\)
is found that is not stable when all component response rates are set to
\(\gamma = 1\), but is stable when half of component response rates are
\(1.95\) and half are \(0.05\). The function is used to illustrate the
concept of how fast versus slow component responses can cause a system
to become stable. Simulations were run for \texttt{iter\ =\ 1000000},
but terminated once an acceptable \texttt{A0} and \texttt{A1} were
found. The code below plots the eigenvalue distributions of \texttt{A0}
and \texttt{A1} in panels \textbf{a} and \textbf{b}, respectively. The
plot itself can be recreated with the function and code below.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{A0 <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(A0[,-}\DecValTok{1}\NormalTok{]);}
\NormalTok{A1 <-}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(A1[,-}\DecValTok{1}\NormalTok{]);}
\NormalTok{plot_Fig_1 <-}\StringTok{ }\NormalTok{function(A0, A1)\{}
    \NormalTok{S_val       <-}\StringTok{ }\KeywordTok{dim}\NormalTok{(A0)[}\DecValTok{1}\NormalTok{];}
    \NormalTok{A0_e        <-}\StringTok{ }\KeywordTok{eigen}\NormalTok{(A0)$values;}
    \NormalTok{A0_r        <-}\StringTok{ }\KeywordTok{Re}\NormalTok{(A0_e);}
    \NormalTok{A0_i        <-}\StringTok{ }\KeywordTok{Im}\NormalTok{(A0_e);}
    \NormalTok{A1_e        <-}\StringTok{ }\KeywordTok{eigen}\NormalTok{(A1)$values;}
    \NormalTok{A1_r        <-}\StringTok{ }\KeywordTok{Re}\NormalTok{(A1_e);}
    \NormalTok{A1_i        <-}\StringTok{ }\KeywordTok{Im}\NormalTok{(A1_e);}
    \NormalTok{A0_vm       <-}\StringTok{ }\NormalTok{A0;}
    \KeywordTok{diag}\NormalTok{(A0_vm) <-}\StringTok{ }\OtherTok{NA}\NormalTok{;}
    \NormalTok{A0vec       <-}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{t}\NormalTok{(A0_vm));}
    \NormalTok{A0vec       <-}\StringTok{ }\NormalTok{A0vec[}\KeywordTok{is.na}\NormalTok{(A0vec) ==}\StringTok{ }\OtherTok{FALSE}\NormalTok{];}
    \NormalTok{A1_vm       <-}\StringTok{ }\NormalTok{A1;}
    \KeywordTok{diag}\NormalTok{(A1_vm) <-}\StringTok{ }\OtherTok{NA}\NormalTok{;}
    \NormalTok{A1vec       <-}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{t}\NormalTok{(A1_vm));}
    \NormalTok{A1vec       <-}\StringTok{ }\NormalTok{A1vec[}\KeywordTok{is.na}\NormalTok{(A1vec) ==}\StringTok{ }\OtherTok{FALSE}\NormalTok{];}
    \NormalTok{fhalf       <-}\StringTok{ }\DecValTok{1}\NormalTok{:(}\FloatTok{0.5}\NormalTok{*}\KeywordTok{length}\NormalTok{(A1vec));}
    \NormalTok{shalf       <-}\StringTok{ }\NormalTok{(}\FloatTok{0.5}\NormalTok{*}\KeywordTok{length}\NormalTok{(A1vec)+}\DecValTok{1}\NormalTok{):}\KeywordTok{length}\NormalTok{(A1vec);}
    \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{mar =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{), }\DataTypeTok{oma =} \KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{));}
    \KeywordTok{plot}\NormalTok{(A0_r, A0_i, }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\FloatTok{3.7}\NormalTok{, }\FloatTok{0.3}\NormalTok{), }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{,}
         \DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{cex.lab =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{cex.axis =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{asp =} \DecValTok{1}\NormalTok{);}
    \NormalTok{vl   <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{0}\NormalTok{, }\DataTypeTok{to =} \DecValTok{2}\NormalTok{*pi, }\DataTypeTok{by =} \FloatTok{0.001}\NormalTok{);}
    \NormalTok{A0x0 <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(S_val) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A0vec) *}\StringTok{ }\KeywordTok{cos}\NormalTok{(vl) +}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{diag}\NormalTok{(A0));}
    \NormalTok{A0y0 <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(S_val) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A0vec) *}\StringTok{ }\KeywordTok{sin}\NormalTok{(vl);}
    \KeywordTok{text}\NormalTok{(}\DataTypeTok{x =} \NormalTok{-}\FloatTok{3.5}\NormalTok{, }\DataTypeTok{y =} \FloatTok{2.25}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"a"}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{);}
    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{A0x0, }\DataTypeTok{y =} \NormalTok{A0y0, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"grey"}\NormalTok{);}
    \KeywordTok{points}\NormalTok{(A0_r, A0_i, }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{);}
    \KeywordTok{plot}\NormalTok{(A1_r, A1_i, }\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\FloatTok{3.7}\NormalTok{, }\FloatTok{0.3}\NormalTok{), }\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{2}\NormalTok{, }\DecValTok{2}\NormalTok{), }\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{,}
         \DataTypeTok{xlab =} \StringTok{""}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }\DataTypeTok{cex.lab =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{cex.axis =} \FloatTok{1.5}\NormalTok{, }\DataTypeTok{asp =} \DecValTok{1}\NormalTok{, }
         \DataTypeTok{col =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{yaxt =} \StringTok{"n"}\NormalTok{);}
    \NormalTok{vl <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \DecValTok{0}\NormalTok{, }\DataTypeTok{to =} \DecValTok{2}\NormalTok{*pi, }\DataTypeTok{by =} \FloatTok{0.001}\NormalTok{);}
    \NormalTok{A0x1a <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(}\FloatTok{0.5}\NormalTok{*S_val) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A1vec[fhalf]) *}\StringTok{ }\KeywordTok{cos}\NormalTok{(vl) +}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{diag}\NormalTok{(A1)[}\DecValTok{1}\NormalTok{:(}\FloatTok{0.5}\NormalTok{*S_val)]);}
    \NormalTok{A0y1a <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(S_val) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A1vec[fhalf]) *}\StringTok{ }\KeywordTok{sin}\NormalTok{(vl);}
    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{A0x1a, }\DataTypeTok{y =} \NormalTok{A0y1a, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"grey"}\NormalTok{);}
    \NormalTok{A0x1b <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(}\FloatTok{0.5}\NormalTok{*S_val) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A1vec[shalf]) *}\StringTok{ }\KeywordTok{cos}\NormalTok{(vl) +}\StringTok{ }
\StringTok{                  }\KeywordTok{mean}\NormalTok{( }\KeywordTok{diag}\NormalTok{(A1)[( (}\FloatTok{0.5}\NormalTok{*S_val) +}\StringTok{ }\DecValTok{1} \NormalTok{):S_val] );}
    \NormalTok{A0y1b <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(}\FloatTok{0.5}\NormalTok{*S_val) *}\StringTok{ }\KeywordTok{sd}\NormalTok{(A1vec[shalf]) *}\StringTok{ }\KeywordTok{sin}\NormalTok{(vl);}
    \KeywordTok{points}\NormalTok{(}\DataTypeTok{x =} \NormalTok{A0x1b, }\DataTypeTok{y =} \NormalTok{A0y1b, }\DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{3}\NormalTok{, }\DataTypeTok{col =} \StringTok{"grey"}\NormalTok{);}
    \KeywordTok{points}\NormalTok{(A1_r[}\DecValTok{1}\NormalTok{:S_val], A1_i[}\DecValTok{1}\NormalTok{:S_val],}\DataTypeTok{pch =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.7}\NormalTok{);   }
    \KeywordTok{text}\NormalTok{(}\DataTypeTok{x =} \NormalTok{-}\FloatTok{3.5}\NormalTok{, }\DataTypeTok{y =} \FloatTok{2.25}\NormalTok{, }\DataTypeTok{labels =} \StringTok{"b"}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{);}
    \KeywordTok{mtext}\NormalTok{(}\DataTypeTok{side =} \DecValTok{1}\NormalTok{, }\StringTok{"Real"}\NormalTok{, }\DataTypeTok{outer =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{line =} \DecValTok{3}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{);}
    \KeywordTok{mtext}\NormalTok{(}\DataTypeTok{side =} \DecValTok{2}\NormalTok{, }\StringTok{"Imaginary"}\NormalTok{, }\DataTypeTok{outer =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{line =} \FloatTok{2.5}\NormalTok{, }\DataTypeTok{cex =} \DecValTok{2}\NormalTok{);}
\NormalTok{\}}
\KeywordTok{plot_Fig_1}\NormalTok{(}\DataTypeTok{A0 =} \NormalTok{A0, }\DataTypeTok{A1 =} \NormalTok{A1);}
\end{Highlighting}
\end{Shaded}

\includegraphics{unnamed-chunk-4-1.pdf}

To find out how frequently \(M\) was stable given that all
\(\gamma = 1\) versus \(\gamma = \{1.95, 0.05\}\), the function below
was created.

>>>>

May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
stability criterion \(\sigma\sqrt{SC} < 1\) assumes that the expected
response rates (\(\gamma\)) of individual components to perturbations of
the system are identical, but this is highly unlikely in any complex
system. In ecological communities, for example, the rate at which
population density changes following perturbation will depend on the
generation time of organisms, which might vary by orders of magnitude
among species. Species with short generation times will respond quickly
(high \(\gamma\)) to perturbations relative to species with long
generation times (low \(\gamma\)). Similarly, the speed at which
individual banks respond to perturbations in financial networks, or
individuals or institutions respond to perturbations in complex social
networks, is likely to vary. The effect of such variance on stability
has not been investigated in complex systems theory. Intuitively,
variation in \(\gamma\) might be expected to decrease system stability
by introducing a new source of variation into the system and thereby
increasing \(\sigma\). Here I show why, despite higher \(\sigma\),
realistic complex systems (such that \(S\) is high but finite) are actually 
more likely to be stable if their individual component response rates vary.

@@

My focus is distinct from Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}}, who applied
the same mathematical framework to investigate how a diagonal matrix
\(\mathbf{X}\) (equivalent to \(\gamma\) in my model) affects the
stability of a community matrix \(\mathbf{M}\) given an interaction
matrix \(\mathbf{A}\) within a generalised Lotka-Volterra model, where
\(\mathbf{M} = \mathbf{XA}\). Gibbs et
al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{17}} analytically
demonstrated that the effect of \(\mathbf{X}\) on system stability
decreases exponentially as system size becomes arbitrarily large
(\(S \to \infty\)). My numerical results do not contradict this
prediction because I did not scale \(\sigma = 1 / \sqrt{S}\), but
instead fixed \(\sigma\) and increased \(S\) to thereby increase total
system complexity, \(\sigma\sqrt{SC}\) (see Supplementary Information for
results simulated across \(\sigma\) and \(C\)). Overall, I show that
component response rate variation increases the upper bound of
complexity at which stability can be realistically observed, meaning
that highly complex systems are more likely than not to vary in their
component response rates, and for this variation to be critical for
system stability.

@@

In the above, \(\mathbf{n^{*}}\) is the vector of species densities at
equilibrium. Feasibility is satisfied if all elements in
\(\mathbf{n^{*}}\) are positive. The matrix \(\mathbf{I}\) is the
identity matrix, and the value \(\theta\) is the strength of
intraspecific competition (diagonal elements). Diagonal values are set
to \(-1\), so \(\theta = -1\). The variable \(\delta\) is a
normalisation parameter that modulates the strength of interactions
(\(\sigma\)) for \(\mathbf{J}\). Implicitly, here \(\delta = 0\)
underlying strong interactions. Hence, \((CS)^{-\delta} = 1\), so in the
above, a diagonal matrix of -1s (\(\theta \mathbf{I}\)) is added to
\(\mathbf{J}\), which has a diagonal of all zeros and an off-diagonal
affecting species interactions (i.e., the expression \((CS)^{-\delta}\)
relates to May's\textsuperscript{\protect\hyperlink{ref-May1972}{1}}
stability
criterion\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}} by
\(\frac{\sigma}{(CS)^{-\delta}}\sqrt{SC} < 1\), and hence for my
purposes \((CS)^{-\delta} = 1\)). Given
\(\mathbf{A} = \theta\mathbf{I + J}\), the above criteria is therefore
reduced to the below\textsuperscript{\protect\hyperlink{ref-Servan2018}{22}},

@@

\hypertarget{ref-Gibbs2017}{}
17. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).

@@

@@

\textbf{This Supplementary information supports the manuscript
``Component response rate variation drives stability in large complex
systems'' with additional analyses to support its conclusions. All text,
code, and data underlying this manuscript are publicly available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub} as
part of the RandomMatrixStability R package.}

The
\href{https://github.com/bradduthie/RandomMatrixStability}{RandomMatrixStability
package} includes all functions and tools for recreating the text, this
Supplementary information, and running all code; additional documentation
is also provided for package functions. The RandomMatrixStability
package is available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub}; to
download it, the
\href{https://cran.r-project.org/web/packages/devtools/index.html}{\texttt{devtools}
library} is needed.

@@

@@

\section{Supplementary Information table of
contents}\label{Supplementary-information-table-of-contents}

\begin{itemize}
\tightlist
\item
  \protect\hyperlink{IncrS}{Stability across increasing \(S\)}
\item
  \protect\hyperlink{ga}{Stability given targeted manipulation of
  \(\gamma\) (genetic algorithm)}
\item
  \protect\hyperlink{ecological}{Stability of ecological networks}

  \begin{itemize}
  \tightlist
  \item
    \protect\hyperlink{competition}{Competitor networks}
  \item
    \protect\hyperlink{mutualism}{Mutualist networks}
  \item
    \protect\hyperlink{pred-prey}{Predator-prey networks}
  \end{itemize}
\item
  \protect\hyperlink{connectance}{Sensitivity of connectance (C) values}

  \begin{itemize}
  \tightlist
  \item
    \protect\hyperlink{connect3}{C = 0.3}
  \item
    \protect\hyperlink{connect5}{C = 0.5}
  \item
    \protect\hyperlink{connect7}{C = 0.7}
  \item
    \protect\hyperlink{connect9}{C = 0.9}
  \end{itemize}
\item
  \protect\hyperlink{sigma}{Large networks of \(C = 0.05\) across \(S\)
  and \(\sigma\)}

  \begin{itemize}
  \tightlist
  \item
    \protect\hyperlink{sigma3}{\(\sigma\) = 0.3}
  \item
    \protect\hyperlink{sigma4}{\(\sigma\) = 0.4}
  \item
    \protect\hyperlink{sigma5}{\(\sigma\) = 0.5}
  \item
    \protect\hyperlink{sigma6}{\(\sigma\) = 0.6}
  \end{itemize}
\item
  \protect\hyperlink{gam_dist}{Sensitivity of distribution of
  \(\gamma\)}
\item
  \protect\hyperlink{Feasibility}{Feasibility of complex systems}
\item
  \protect\hyperlink{Gibbs}{Consistency with Gibbs et al. (2018)}
\end{itemize}

### 2019-03-03 notebook/ms.Rmd
Tweaks

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) in which $\gamma$ varies are actually more likely to be stable.

>>>>

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that the expected response rates ($\gamma$) of individual components to perturbations of the system are identical, but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of organisms, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance on stability has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, realistic complex systems (such that $S$ is high but finite) are actually more likely to be stable if their individual component response rates vary
